---
title: "Machine Learning 1 - Earthquakes"
author: "Esin, Andrea and Sabrina"
date: "2023-06-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Requirements group work:
 - write a nice story
 - compare methods
 - Conclusion
 
 Length: no more than 30 pages
 Deadline: 9 June


## Significant Earthquakes

We are looking at the significant earthquakes from 2150 B.C. to 2019. Significant means that at least one of these criteria was fulfilled:

- Damage 1 Million Dollar or more

- 10 or more deaths

- Magnitude of 7.5 or more

- Modified Mercalli Intensity X or greater

- Earthquake generated a tsunami


```{r data, message=FALSE, warning=FALSE, include=FALSE}

setwd("C:/Users/A/Documents/FS23_ML1/Project Data/Git_repository/ML01")
eqdata <- read.csv("significant-earthquake-database.csv", header = TRUE, sep = ";")

head(eqdata)
mean(eqdata$Intensity, na.rm = TRUE)


```
At first we will have a look at the data provided.

```{r overview}
head(eqdata)
summary(eqdata)
mean(eqdata$Intensity, na.rm = TRUE)
str(eqdata)

```



You can also embed plots, for example:

```{r intensity vs deaths, echo=FALSE}

plot(eqdata$Earthquake...Deaths ~ eqdata$Intensity,
     main = "Intensity against Deaths")

```


